Histogram of Gradient Orientations of Signal Plots applied to P300 Detection

### 2017-12-12 article.tex
comentarios

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother 
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2016}
\copyrightyear{2016}
\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{Oriented Gradient Histogram applied to P300 Detection}

% Authors, for the paper (add full first names)
\Author{Rodrigo Ramele $^{1,\dagger}$*, Ana Julia Villar $^{1}$ and Juan Miguel Santos  $^{1}$}
% Authors, for metadata in PDF
\AuthorNames{Rodrigo Ramele, Ana Julia Villar and Juan Miguel Santos}

@@

% Current address and/or shared authorship
\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina}

@@

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The analysis of Electroencephalographic (EEG) signals is of ulterior importance to elucidate patterns that could improve the implementation of Brain Computer Interfaces (BCI). These systems are meant to provide alternative pathways to transmit volitional information which could potentially enhance the quality of life of patients affected by neurodegenerative disorders or improve Human Computer Interaction systems.  Of particular interests are those which are based on the recognition of Event-Related Potentials (ERP) because they can be elicited by external stimuli and used to implement spellers, to control external devices or even avatars in virtual reality environments.  This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from the images of the plots of the signals.  It also aims to provide a framework to analyze, characterize and classify EEG signals, with a focus on the P300, an ERP elicited by the oddball paradigm of rare events.  The validity of the method is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients.}

% Keywords
\keyword{electroencephalography (EEG); BCI; P300; ALS; NBNN; HOG; SIFT}

@@

%\begin{enumerate}[leftmargin=*,labelsep=3mm]
%\item	Why
%\item	Purpose and significance
%\item	Current state of the research field and key publications cited
%\item   Controversial hypothesis
%\item   Main aim and principal conclusions.
%\end{enumerate}
%
%\begin{enumerate}[leftmargin=*,labelsep=3mm]
%\item El paper tiene un foco general en EEG y rapidamente se centra en BCI como una prueba del método.
%\item La relevancia de estudiar métodos de análisis de EEG para pacientes con ALS.
%\item Ofrece una Receta de cómo implementar un esquema simple de detección de P300 y ofrece una comparativa contra el paper de las italianas.
%\item La transversalidad del método de poder detectar eventos rítmicos asi como transientes como p300.
%\item La posibilidad de utilizarlo para reconocer, fuertemente un patron particular y con la flexibilidad de detectar variantes basado en las poblaciones de descriptores (esta es mi hipótesis de porque funciona, porque yo vi que el método se hace pelota con pequeñas variantes de la señal).
%\item La importancia del averaging de EEG y sus desbalanceos, puntualizando que en el paper original
%\item Que la caída en la clasificación no cae significativamente aún bajando la cantidad de repeticiones requeridas para el averaging.
%\item ¿ Por qué da mejor en PO8 que en Cz y Pz que son los canales P300 ?  (como future work).
%\end{enumerate}

Although recent advances in neuroimagining techniques (particularly radio-nuclear and radiological scanning methods) \citep{Schomer2010} have diminished the prospects of the traditional Electroencephalography (EEG), the advent and development of digitalized devices has pressed for a revamping of this hundred years old technology.  Their versatility, ease of use, temporal resolution, ease of development and fabrication, and its proliferation as consumer devices, are pushing EEG to become the de-facto non invasive portable or ambulatory method to access and harness brain information \cite{DeVos2014}

A key contribution to this expansion has been the field of Brain Computer Interfaces (BCI) \citep{WolpawJonathanR2012} which is the pursuit of the development of a new channel of communication particularly aimed to persons affected by neurodegenerative diseases.

One noteworthy aspect of this novel communication channel is the ability to volitionally transmit information from the Central Nervous System (CNS) to a computer device and from there use that information to control a wheelchair \citep{Carlson2013}, as input to a speller application \citep{Guger2009a}, in a Virtual Reality environment \citep{Lotte2013} or as aiding tool in a rehabilitation procedure \citep{Jure2016}.  The holly grail of BCI is to implement a new complete and alternative pathway to restore lost locomotion \citep{WolpawJonathanR2012}.

@@

Moreover, EEG markers \citep{Clerc} that can be used to volitionally transmit information are limited, and each one of them has a particular combination of appropriate methods to decode them. Inevitably, it is necessary to implement  distinct and specialized algorithmic methods, to filter the signal, enhance its Signal to Noise Ratio (SNR), and try to determine some meaning out of it.

%Patients suffering from ALS their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns alterations \citep{Nijboer2009,Riener2014}.

%Besides novel applications of BCI, the goal of the entire discipline is to provide communication assistance to people affected by neuro-degenerative diseases.  It is understood that for people suffering from ALS, even though this pathology do not harm the CNS, their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns alterations \citep{Nijboer2009,Riener2014}.

BCI has gained mainstream public awareness with worldwide challenge competitions like Cybathlon  \citep{Riener2014} and even been broadcasted during the inauguration ceremony of the 2014 Soccer World Cup.  New developments have overcome the out-of-the-lab high-bar and they are starting to be used in real world environments \citep{Huggins2016}.  However, they still lack the necessary robustness, and its performance is well behind any other method of human computer interaction, including any kind of detection of residual muscular movement~\citep{Clerc}.

In~\citep{Ramele2016},  authors introduce a method for classification of rhythmic patterns using gradient histograms orientations. Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events like those produced by P300 Event Related Potential. 
 Its validity is verified by processing offline data for ALS patients.  
The proposed approach is based on the morphological analysis of the shape of the EEG signal~\citep{Alvarado-Gonzalez2016,Yamaguchi2009} and it was built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting  raw signal plots~\citep{Hartman2005}.

This paper reports a method to, (1) classify P300 signals based on the identification of their structure in the shape domain using histograms of oriented gradients extracted from the image of signal plots, and (2) describe the way in which this classification can be used to implement an offline P300-based BCI Speller application using two public datasets.

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradients of the Signal Plot method is explained. Section \ref{Preprocessing} and \ref{Pipeline} describe the processing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008} and the final Section shows results and discussion where we expose our remarks, conclusions and future work.

%\subsection{Morphological Signal Analysis}
%
%Several approaches and methods have been applied to decode them.  Of particular interests are morphological signal analysis 
%CITE (Paper del chabon del ITBA que detalla bien los diferentes metodos) \citep{Alvarado-Gonzalez2016,Yamaguchi2009}.
%
%Morphological or 
%
%template matching
%
%spike
%sharp wave
%spike wave
%polyspike 
%slow wave complex
%
%shape domain
%
%\subsection{P300}
%
%Uhhhh I can talk a lot about P300 \citep{Knuth2006}.
%
%P300 single trial is a golden grail of Brain computer interfaces.  It is studied and analyzed in this way, here in this other way, in this other way is here.

@@

\subsection{Feature Extraction based on Histogram of Gradients of the Signal Plot} \label{Feature}

The P300~\citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs around $300$ ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm~\cite{WolpawJonathanR2012} and it is consistent across different subjects. It has a lower amplitude  ($\pm 5 \mu V $) compared to basal EEG activity, reaching a SNR of around $-15$ db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity~\citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix~\citep{Farwell1988}. Fig.~\ref{fig:p300matrix} shows an example of the Speller Matrix used in the OpenVibe Open Source software~\citep{Renard2010}, where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

\begin{figure}[H]
\centering
\subfigure[Row flash.]{\includegraphics[width=.29\linewidth]{openvibep300matrix.png}}
\subfigure[Column flash.]{\includegraphics[width=.29\linewidth]{p300matrixcoloriginal.png}}
\caption{Example of a Speller Matrix.  Rows and columns flash intermittently in random permutations.}
\label{fig:p300matrix}
\end{figure}

%The proposed algorithm was designed to perform specifically this decoding.

% of the selected letter based on the detection of the P300 Event Related Potential response on the corresponding EEG signal.

\subsubsection{Preprocessing} \label{Preprocessing}

The preprocessing stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. 
The processing pipeline starts by applying a notch filter to the raw signal, a 
$4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \citep{Krusienski2006}.

\subsubsection{Processing Pipeline} \label{Pipeline}

\begin{itemize}
\item \textbf{Artefact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the $6$ rows plus the $6$ columns, a basic artefact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ stimuli,  a window of $1$ second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments are labeled as \textit{hit}, whereas the remaining $10$ are labeled as \textit{no hit}.  A hit represents that the EEG segment should contain the P300 ERP signature time-locked to the flashing stimulus. 
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out \citep{Liang2008}. 
\end{itemize}

@@

The underlying idea of this method is to generate a template for the signal \citep{Ramele2016},  based on the image plot.  Hence, the first step is the signal transformation into a temporary binary image.

Averaged signal segments obtained from the preprocessing stage are first scaled and standardized (i.e. z-score) by

\begin{equation}
\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot \frac{( x(t,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor
\end{equation}

\noindent where $\gamma$ is the image scale, $t$ is time and $ x(t,c) $ is the point-to-point averaged EEG matrix for time $t$ and for  channel $c$. Lastly, $\bar{x}(c)$ and $ \sigma (c) $ are the mean and standard deviation of $x$.

@@

\begin{equation}
I(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \cdot t; z_2 = \tilde{x}(t,c) + z(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

\noindent where $ z_1$ and $z_2$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image.  The function $z(c)$ is the \textit{zerolevel} which is the location on the image where the signal's zero value should be located in order to fit the entire signal within the image:

\begin{equation}
z(c) = \left \lfloor{ \frac{\max \tilde{x}(t,c)  - \min \tilde{x}(t,c) }{2} }\right \rfloor -   \left \lfloor{ \frac{\max \tilde{x}(t,c)  + \min \tilde{x}(t,c)}{ 2} }\right \rfloor
\label{eq:zerolevel}
\end{equation}

@@

%\noindent and is used to derive local representations that captures the visual shape of the signal.

\subsubsection{Feature Extraction: Histogram of Oriented Gradients}
\label{SIFT}

@@

On the generated image, a keypoint $\mathbf{kp}$ is placed on a pixel $(x_{kp}, y_{kp})$ over the image plot and a window around the keypoint is considered. A local image patch of size $S_p \times S_p$ is constructed by dividing the window in $16$ blocks of size $3s$ each one,  where $s$ is the scale of the local patch and it is an input parameter of the algorithm. It is arranged in a $4 \times 4$ grid and the pixel $ \mathbf{kp}$ is the patch center, then $S_p = 4.3s $ pixels.

@@

\noindent  where $ \theta \in \{0, 45, 90, 135, 180, 225, 270, 315\} $, $ i,j \in \{1,2,3,4\} $, $ |J(\mathbf{x})| $ is the norm of the gradient vector found at each block of the patch using finite differences, $\angle J(\mathbf{x}) $ is the angle of the gradient vector,  $\theta$ is the angle bin, $\mathbf{x}$ is a pixel from  the $i,j$-block $B_{i,j}$ and $ w_\mathrm{ang}(\cdot) $  and $ w_{ij}(\cdot) $ are linear interpolation functions used by Lowe and Vedaldi et al. in~\citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each block when $s = 1$.  As the patch has  $16$ blocks and  $8$ bin angles are considered, a descriptor of $128$ dimension is obtained. It can be observed that in each step, the histogram is computed by multiplying by $ |J(\mathbf{x})| $ , so the method considers both, the magnitude and the orientation of the gradient vector.

@@

The aim is to identify the selected letter from the matrix. Previously, during training phase, descriptors from \textit{hit} labeled segments are extracted.  These descriptors are the P300 templates which are grouped in a template set called $ T $.  Segments corresponding to rows are labeled 1-6, whereas those corresponding to columns are labeled 7-12.  The process has the following steps:

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item Repeat item 1 $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t$ and $c$ correspond to time and chanel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item Compute  
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c)
\label{averaging}
\end{equation}

\item Plot the signal $x(t,c)$, according Section~\ref{Plot}.

\item Obtain the descriptor from the image of the signal plot of $x(t,c)$ in accordance to the method described in Section~\ref{SIFT}, $ d^r_1, \dots,  d^r_6 $ for rows and  $ d^c_7, \dots,  d^c_{12} $ for columns.

@@

\begin{equation}
\hat{r} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q \in NN_T(d^r_u)}^{} \left\lVert q -  d^r_u \right\rVert ^2
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
\hat{c} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q \in NN_T(d^c_u)}^{} \left\lVert q -  d^c_u \right\rVert ^2
\label{eq:multiclassificationcol}
\end{equation}

where $NN_T(d^l_u), \;l=r,c$ is the set of the $k$ nearest neighbors to $d^l_u, \;l=r,c$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l=r,c$.  By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{r} $ and column $ \hat{c} $.

@@

\end{enumerate}

Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}
\centering
\includegraphics[width=15cm]{classificationgraph2.pdf}
\caption{Single trial segments are averaged for the 6 rows and 6 columns.}
\label{fig:classification}
\end{figure}

%In the case of the P300 response, the oddball paradigm requires that one of the stimuli be infrequent. Hence this forces the data to be unbalanced~\citep{Tibon2015}.  At the same time, the NBNN method suffers from biased classification on unbalanced classes~\citep{Fornoni2014}. %Para solucionar este problema,

@@

%------------------
\subsection{Experimental Protocol} \label{Protocol}

@@

We replicated the same experiment on healthy subjects using a wireless digital EEG device (g.Nautilus, g.Tec, Austria).  The experimental conditions were the same as those used for the previous dataset, as detailed in section~\ref{ALSDataset}.

Participants were recruited voluntarily and the experiment were conducted anonymously in accordance with the declaration of Helsinki published by the World Health Organization.  No monetary compensation was handed out and all participants agreed and signed a written informed consent.  All healthy subjects had normal or corrected-to normal vision and no history of neurological disorders. The experiment was performed with 20 subjects, 7 males, 4 females, average age 22.5 years, standard deviation  1,8 years, range 21–26 years.

Gel-based active electrodes (g.LADYbird, g.Tec, Austria) were used on the same locations Fz, Cz, P3, Pz, P4, PO7, PO8 and Oz.  Reference was set to the right ear lobe and ground was preset as the AFz position.   Sampling frequency is slightly different, and was set to 250 Hz, which is the closest possible to the one used with the other dataset.

>>>>

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother 
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2017}
\copyrightyear{2017}
\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{Histogram of Gradient Orientations of Signal Plots applied to P300 Detection}

% Authors, for the paper (add full first names)
\Author{Rodrigo Ramele $^{1}$*, Ana Julia Villar $^{1}$ and Juan Miguel Santos  $^{1}$}
% Authors, for metadata in PDF
\AuthorNames{Rodrigo Ramele, Ana Julia Villar and Juan Miguel Santos}

@@

% Current address and/or shared authorship  COMMENTED PER REVIEWER
%\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina}

%\firstnote{.}

@@

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The analysis of Electroencephalographic (EEG) signals is of ulterior importance to elucidate patterns that could improve the implementation of Brain Computer Interfaces (BCI). These systems are meant to provide alternative pathways to transmit volitional information which could potentially enhance the quality of life of patients affected by neurodegenerative disorders or improve Human Computer Interaction systems.  Of particular interests are those which are based on the recognition of Event-Related Potentials (ERP) because they can be elicited by external stimuli and used to implement spellers, to control external devices or even avatars in virtual reality environments.  This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from the images of the plots of the signals.  It also aims to provide a framework to analyze, characterize and classify EEG signals, with a focus on the P300, an ERP elicited by the oddball paradigm of rare events.  The validity of the method is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients and an own dataset for healthy subjects.}

% Keywords
\keyword{EEG; BCI; P300; ALS; NBNN; HGOSP; SIFT}

@@

Although recent advances in neuroimagining techniques (particularly radio-nuclear and radiological scanning methods) \citep{Schomer2010} have diminished the prospects of the traditional Electroencephalography (EEG), the advent and development of digitized devices has pressed for a revamping of this hundred years old technology.  Their versatility, ease of use, temporal resolution, ease of development and fabrication, and its proliferation as consumer devices, are pushing EEG to become the de-facto non invasive portable or ambulatory method to access and harness brain information~\cite{DeVos2014}.

A key contribution to this expansion has been the field of Brain Computer Interfaces (BCI)~\citep{WolpawJonathanR2012} which is the pursuit of the development of a new channel of communication particularly aimed to persons affected by neurodegenerative diseases.

One noteworthy aspect of this novel communication channel is the ability to transmit information from the Central Nervous System (CNS) to a computer device and from there use that information to control a wheelchair~\citep{Carlson2013}, as input to a speller application~\citep{Guger2009a}, in a Virtual Reality environment~\citep{Lotte2013} or as aiding tool in a rehabilitation procedure~\citep{Jure2016}.  The holly grail of BCI is to implement a new complete and alternative pathway to restore lost locomotion~\citep{WolpawJonathanR2012}.

@@

Moreover, EEG markers~\citep{Clerc} that can be used to  transmit volitional information are limited, and each one of them has a particular combination of appropriate methods to decode them. Inevitably, it is necessary to implement  distinct and specialized algorithmic methods, to filter the signal, enhance its Signal to Noise Ratio (SNR), and try to determine some meaning out of it.

BCI has gained mainstream public awareness with worldwide challenge competitions like Cybathlon~\citep{Riener2014} and even been broadcasted during the inauguration ceremony of the 2014 Soccer World Cup.  New developments have overcome the out-of-the-lab high-bar and they are starting to be used in real world environments~\citep{Guger2017,Huggins2016}.  However, they still lack the necessary robustness, and its performance is well behind any other method of human computer interaction, including any kind of detection of residual muscular movement~\citep{Clerc}.

A few works have explored the idea of exploiting the signal waveform to analyze the EEG signal.  In~\citep{Alvarado-Gonzalez2016} an approach based on Slope Horizontal Chain Code is presented, whereas in ~\citep{Yamaguchi2009} a similar procedure was implemented based on Mathematical Morphological Analysis.  The seminal work of Bandt-Pompe Permutation Entropy~\citep{Berger2017} also explores succinctly this idea as a basis to establish the time series ordinal patterns.  In the article~\citep{Ramele2016},  the authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves  and Motor Imagery rolandic central $\mu$ rhythms using the histogram of gradient orientations of signal plots.  Inspired in that work, we propose a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  
The proposed approach is based on the waveform analysis of the shape of the EEG signal, but using histogram of gradient orientations. The method is built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting raw signal plots, as it is described in~\citep{Hartman2005}.

This paper reports a method to, (1) classify P300 signals based on the identification of their structure in the shape domain using histograms of gradient orientations extracted from the image of signal plots, and (2) describe the way in which this classification can be used to implement an offline P300-based BCI Speller application using two public datasets. Its validity is verified by processing offline data for ALS patients and for healthy subjects.

This article unfolds as follows: Section~\ref{Feature} is dedicated to explain the Feature Extraction method based on Histogram of Gradient Orientations of the Signal Plot: Section~\ref{Pipeline} shows the preprocessing pipeline,  Section~\ref{Plot}  describes the image generation of the signal plot, Section~\ref{SIFT}  presents the feature extraction procedure while  Section~\ref{Classification}  introduces the whole  Speller Matrix Letter Identification procedure including the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008}. In Section~\ref{Protocol}, the experimental protocol is expounded. Section~\ref{Results} shows the results of applying the proposal technique and a discussion is presented. 
In the final Section~\ref{conclusion}  we expose our remarks, conclusions and future work.
%This kind of EEG events are described in~\citep{Hartman2005} and~\citep{Ben-Simon2013}.

@@

The P300~\citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs around $300$ ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm~\cite{WolpawJonathanR2012} and it is consistent across different subjects. It has a lower amplitude  ($\pm 5 \mu V $) compared to basal EEG activity, reaching a SNR of around $-15$ db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity~\citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix~\citep{Farwell1988}. Fig.~\ref{fig:p300matrix} shows an example of the Speller Matrix used in the OpenVibe Open Source software~\citep{Renard2010}, where the flashes of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or a column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

\begin{figure}[H]
\centering
\subfigure[Row flash.]{\includegraphics[width=.29\linewidth]{openvibep300matrix.png}}
\subfigure[Column flash.]{\includegraphics[width=.29\linewidth]{p300matrixcoloriginal.png}}
\caption{Example of the $6 \times 6$ Speller Matrix used in the study.  Rows and columns flash intermittently in random permutations.}
\label{fig:p300matrix}
\end{figure}

\subsection{Feature Extraction based on Histogram of Gradient Orientations of the Signal Plot} \label{Feature}

In this section, the signal preprocessing, the method for generating images from signal plots, the feature extraction procedure and the speller matrix identification are described.

\subsubsection{Preprocessing Pipeline} \label{Pipeline}

\begin{itemize}
\item \textbf{Signal Enhancement}: The preprocessing stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. The processing pipeline starts by applying a notch filter to the raw digital signal, a 
$4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \citep{Krusienski2006}.
\item \textbf{Artifact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the $6$ rows plus the $6$ columns, a basic artefact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ stimuli,  a window of $t_{max} = 1$ second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments should contain the P300 ERP signature time-locked to the flashing stimulus, one for the row, and one for the column.
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out~\citep{Liang2008}. 
\end{itemize}

@@

Averaged signal segments are standardized and scaled by

\begin{equation}
\tilde{x}(n,c) = \left \lfloor{ \gamma \cdot \frac{( x(n,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor, n \in \{ 1, \dots, n_{max}\},c \in \{1,2,\dots,Ch\}
\label{eq:standarizedaverages}
\end{equation}

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related with the image scale. In addition, $ x(n,c) $ is the point-to-point averaged EEG matrix for the sample point $n$ and for channel $c$, $n_{max}=F_s.t_{max}$ and $F_s$ is the sampling frequency. The parameter  $Ch$ is the number of available EEG channels. Lastly, $$\bar{x}(c) =\frac{1}{n_{max}}\sum_{n=1}^{n_{max}}x(n,c)$$ and $$ \sigma(c) = (\frac{1}{n_{max}-1}\sum_{n=1}^{n_{max}}(x(n,c)-\bar{x}(c))^2 )^{\frac{1}{2}}$$ are the mean and standard deviation of $x(n,c), n \in 1,\dots,n_{max}$, for each channel $c$.

@@

\begin{equation}
I(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \cdot n; \; z_2 = \tilde{x}(n,c) + z(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

\noindent where $ (z_1,z_2) \in \mathbb{N} \times \mathbb{N}$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image,  $n \in \{1, \dots, n_{max}\}$ and $c \in \{1,2,\dots,Ch\}$.  The function $z(c)$, $c \in \{1,2,\dots,Ch\}$ is the location on the image where the signal's zero value has to be located in order to fit the entire signal within the image:

\begin{equation}
z(c) = \left \lfloor{ \frac{\max_{1 \leq n\leq n_{max}} \tilde{x}(n,c)  - \min_{1 \leq n\leq n_{max}} \tilde{x}(n,c) }{2} }\right \rfloor -   \left \lfloor{ \frac{\max_{1 \leq n\leq n_{max}} \tilde{x}(t,c)  + \min_{1 \leq n\leq n_{max}} \tilde{x}(n,c)}{ 2} }\right \rfloor
\label{eq:zerolevel}
\end{equation}

@@

\subsubsection{Feature Extraction: Histogram of Gradient Orientations}
\label{SIFT}

@@

On the generated image $I$, a keypoint $\mathbf{kp}$ is placed on a pixel $(x_{kp}, y_{kp})$ over the image plot and a window around the keypoint is considered. A local image patch of size $S_p \times S_p$ pixels is constructed by dividing the window in $16$ blocks of size $3s$ each one,  where $s$ is the scale of the local patch and it is an input parameter of the algorithm. It is arranged in a $4 \times 4$ grid and the pixel $ \mathbf{kp}$ is the patch center, thus $S_p = 4.3s $ pixels.

@@

\noindent  where $\mathbf{x}$ is a pixel from  the $i,j$-block $B_{i,j}$, $ i,j \in \{1,2,3,4\} $, $ \theta \in \{0, 45, 90, 135, 180, 225, 270, 315\} $,  $ |J(\mathbf{x})| $ is the norm of the gradient vector in the pixel $\mathbf{x}$ and it is computed using finite differences, $\angle J(\mathbf{x}) $ is the angle of the gradient vector,  $\theta$ is the angle bin,   and $ w_\mathrm{ang}(\cdot) $  and $ w_{ij}(\cdot) $ are linear interpolation functions used by Lowe and Vedaldi et al. in~\citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each block when $s = 1$.  As the patch has  $16$ blocks and  $8$ bin angles are considered, a descriptor of $128$ dimension is obtained. It can be observed that in each step, the histogram is computed by multiplying by $ |J(\mathbf{x})| $, so the method considers both, the magnitude and the orientation of the gradient vector.

@@

The aim is to identify the selected letter from the matrix. Previously, during training phase, two descriptors are extracted from averaged signal segments which correspond to the letter where the user was supposed to be focusing onto.  These descriptors are the P300 templates which are grouped in a template set called $ T $.  Segments corresponding to rows are labeled 1-6, whereas those corresponding to columns are labeled 7-12.  The process has the following steps:

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the single trial segments $S_1(n,c),\dots,S_{k_a}(n,c)$, where the variables $n \in \{1, \dots, n_{max}\}$ and $c \in \{1,2,\dots,Ch\}$ correspond to time and channel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item \label{paso3} Compute the Ensemble Average by
\begin{equation}
x(n,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(n,c),n \in \{1, \dots, n_{max}\}, c \in \{1,\dots,Ch\}
\label{averaging}
\end{equation}  
for each row and column.

\item \label{paso4}Plot the signal $x(n,c)$,  $n \in \{1, \dots, n_{max}\}$, $c \in \{1,\dots,Ch\}$,  according Section~\ref{Plot}.

\item Repeat steps~\ref{paso2}, \ref{paso3} and~\ref{paso4} in order to generate the images $I^{row}_1, \dots, I^{row}_6$ and $I^{col}_7,\dots,I^{col}_{12}$ for rows and columns, respectively.

\item Obtain the descriptors $ d^{row}_1, \dots,  d^{row}_6 $  and  $ d^{col}_7, \dots,  d^{col}_{12} $ for rows and columns, respectively from $I^{row}_1, \dots, I^{row}_6$ and $I^{col}_7,\dots,I^{col}_{12}$  in accordance to the method described in Section~\ref{SIFT}.

@@

\begin{equation}
\hat{row} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q \in NN_T(d^{row}_u)}^{} \left\lVert q -  d^{row}_u \right\rVert ^2
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q \in NN_T(d^{col}_u)}^{} \left\lVert q -  d^{col}_u \right\rVert ^2
\label{eq:multiclassificationcol}
\end{equation}

where $NN_T(d^l_u), \;l\in\{row,{col}\}$ is the set of the $k$ nearest neighbors to $d^l_u, \;l\in\{row,{col}\}$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l\in\{row,{col}\}$.

@@

\end{enumerate}
By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{row} $ and column $ \hat{col} $. 
Figure~\ref{fig:classification} shows a scheme of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments $S_i$ are averaged for the 6 rows and 6 columns. From the averaged signal, the image of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure}

@@

%------------------
%In the case of the P300 response, the oddball paradigm requires that one of the stimuli be infrequent. Hence this forces the data to be unbalanced~\citep{Tibon2015}.  At the same time, the NBNN method suffers from biased classification on unbalanced classes~\citep{Fornoni2014}. %Para solucionar este problema,

\subsection{Experimental Protocol} \label{Protocol}

@@

We replicate the same experiment on healthy subjects using a wireless digital EEG device (g.Nautilus, g.Tec, Austria).  The experimental conditions are the same as those used for the previous dataset, as detailed in section~\ref{ALSDataset}.

Participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and all participants agree and sign a written informed consent.  All healthy subjects have normal or corrected-to normal vision and no history of neurological disorders. The experiment is performed with 8 subjects, 6 males, 2 females, 6 right-handed, 2 left-handed, average age 28.25 years, standard deviation  9.58 years, range 20-50 years.

EEG data is collected in a single recording session. Participants are seated in a comfortable chair, with their vision aligned to a computer screen located one meter in front of them.  The handling and processing of the data and stimulations is conducted by the OpenVibe platform~\citep{Renard2010}.
